Using VariantSpark to learn population structure from genomic profiles

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
logreg results
chr22 results

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
\marginpar{TODO elaborate on the error} 
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to XXX.  We therefore use only chromosome 22 for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 5-fold cross-validation and report the mean accuracy over all training and testing folds
along with the standard error.  We also perform 5 runs of \kMeans{} with 4 clusters and report the mean accuracy of the
resulting models along with the standard error.

\marginpar{TODO elaborate on accuracy measures}
As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately (XX\%, ARI=Y), see table~\ref{RF}.
RF achieves a higher accuracy (XX\%, ARI=Y) but is outperformed by the simple logistic regression (XX\%, ARI=Y).
This indicates that a multivariate approach may not be needed to accurately fit ethnicity.

@@

\marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
  paper)}
\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Bejing, China), and GBR (British in England and
Scotland).

@@

Given the perfect result when building the logistic regression model on this small subset of data, we expand our dataset
to include every population from the 1000 Genomes Project phase1 dataset.

@@

To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR,
EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we cluster
the entirety of these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.  Given the perfect result from the
logistic regression models on the three populations, and the improvement over \kMeans{}, we proceed with logistic
regression on the four super-populations.

>>>>

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
\marginpar{TODO elaborate on the error} 
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and report the mean accuracy over all training and testing folds
along with the standard error.  We also perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean accuracy of the
resulting models along with the standard error.

\marginpar{TODO elaborate on accuracy measures}
As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
Although capable of fitting a relatively accurate model (ARI=0.832), the average ARI over ten models is  \(0.553 \pm 0.07\), see table~\ref{RF}.
RF achieves a better fitting model (XX\%, ARI=Y) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.02\).
This indicates that a multivariate approach may not be needed to accurately fit ethnicity.

@@

\marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
  paper)}
\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether the multivariate approach, deep learning, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Beijing, China), and GBR (British in England and
Scotland).

@@

@@

To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR,
EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we cluster
the entirety of these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.  Given the perfect result when building
the logistic regression model on this small subset of data, we expand our dataset to include every population from the
1000 Genomes Project phase1 dataset.
